Master's Thesis Summary

### 2023-03-06 Master's Thesis Summary/summary.tex
LPAR

\documentclass[a4paper,11pt]{report}

\usepackage{setspace}
\usepackage{caption}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{bussproofs}
\usepackage{enumerate}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage{stackengine}
\usepackage{footmisc}
\usepackage{wrapfig}
\usepackage[a4paper,top=3cm,bottom=3.8cm,left=3cm,right=3cm,headsep=20pt]{geometry}
\usepackage[toc,page]{appendix}
\usepackage{listings}
\lstset{
	language=bash,
	basicstyle=\ttfamily
}

\usepackage{hyperref}% http://ctan.org/pkg/hyperref
\usepackage{cleveref}% http://ctan.org/pkg/cleveref

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\begin{document}

	\pagenumbering{gobble}
	\newgeometry{top=2cm, right=2.5cm, left=2.5cm, bottom=2.5cm}
	\begin{flushright}
		% https://public.univie.ac.at/en/downloads/logos/
		\includegraphics[width=7cm]{RZ_Logo_Uni_sw}
	\end{flushright}	
	\vskip20mm
	\begin{center}
		\huge\textbf{MASTERARBEIT / MASTER'S THESIS}
		\vskip15mm
		\normalsize{Titel der Masterarbeit / Title of the Master's Thesis}

		\LARGE\textbf{``Automated Deduction for Intuitionistic Logic\\via Embedding into Classical Logic''} % Insert title here
		\vskip15mm
		\normalsize{verfasst von / submitted by}

		\Large\textbf{Alexander Pluska, BSc} % Insert your name in this line
		\vskip15mm
		\normalsize{angestrebter akademischer Grad / in partial fulfillment of the requirements for the degree of}

		\Large\textbf{Master of Science (MSc)}

		\small
	\end{center}
	\vskip15mm
	\begingroup
	\renewcommand{\arraystretch}{1.2}
	\begin{flushleft}
		Wien, 2022 / Vienna, 2022 % Insert graduation year in this line
		\vfill
		\addtolength{\tabcolsep}{-6pt} 
		\begin{table}[h!]
			\begingroup
			\renewcommand*{\arraystretch}{1.2}
			%\renewcommand*{\linespacing}{1}
			\renewcommand\baselinestretch{1}
			\small
			\begin{tabular}{p{7cm} p{10cm}}
				Studienkennzahl lt. Studienblatt / \newline degree programme code as it appears on \newline the student record sheet: &UA 066 821 \\
				Studienrichtung lt. Studienblatt / \newline degree programme as it appears on \newline the student record sheet:& Masterstudium Mathematik\\
				Betreut von / Supervisor: & Univ.-Prof. Matthias Aschenbrenner, PhD
			\end{tabular}
			\endgroup
		\end{table}
		\addtolength{\tabcolsep}{6pt} 
	\end{flushleft}	
	\endgroup	
	\restoregeometry

	\newpage
	\
	\newpage

	% Some of the code for creating ToC id adapted from the following template:
	% https://www.sharelatex.com/templates/thesis/university-of-sussex-thesis

		\
		\vfill
		\onehalfspacing
		\begin{center}
		\bfseries Abstract		
		\end{center}
		The famous double negation translation~\cite{glivenko1929quelques, godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously, the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.

		Furthermore, we implement a system that takes intuitionistic first-order problems in the tptp file format as an input and performs our transformation. This allows the use of classical theorem provers for checking intuitionistic validity. We benchmark our implementation using the Vampire theorem prover~\cite{kovacs2013first} on the ILTP problem set~\cite{iltp}.

		Finally, we discuss how the generated classical proofs of the transformed problem can be translated back into intuitionistic proofs of the original problem. All in all, this establishes a novel approach to theorem proving for intuitionistic logic and provides a first proof of concept.

	\vspace*{1cm}
	\begin{center}
		\bfseries Acknowledgements		
	\end{center}

	I want to thank Matthias for accommodating my particular interests by agreeing to supervise a thesis which lies outside his main area of research and bearing with the organizational difficulties.

	I sincerely appreciate Florian's guidance, who graciously agreed to co-supervise a thesis at a different university. He has provided me with a fascinating topic as well as plenty of his time, expertise and thorough feedback.

	Finally, I am incredibly grateful to my family, in particular my parents Stefania and Cezary, who have instilled in me the joy of learning, provided all the love and help I could ask for and invariably supported me along my journey.

	{\vfill\null}

	\newpage
	\
	\singlespacing
	\newgeometry{top=1cm, bottom=2cm}
	\tableofcontents
	\restoregeometry

	\chapter{Introduction}
	\onehalfspacing

	\pagenumbering{arabic}
	Constructive mathematics refers to a flavour of mathematics in which the existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction.
	The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the fact that the law of excluded middle $A\vee\neg A$ and the double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.
	Besides philosophical considerations, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation for studying constructive mathematics from the perspective of computer science in that proofs in  intuitionistic logic directly correspond to computer programs --- as expressed in the Curry-Howard correspondence~\cite{howard1980formulae}.

@@

	\chapter{Overview}

	The main content is organized in chapters 3 to 6. Chapter 3 collects prerequisite definitions and serves to fix our notation and recall some important facts. Chapter 4 is the meat of this thesis. In particular, it features a complete description of our embedding of intuitionistic into classical logic as well as proving its correctness. In chapter 5 we present the implementation of the embedding and the results of our benchmark. Finally, in chapter 6 we explicitly describe how to translate the generated proofs and counter-models which in principle directly follows from the results of section 4. 	We shall now for each chapter give an overview of our main results and highlight the key arguments.

	\section{Embedding intuitionistic into classical logic}

	Recall that our goal in this chapter is to give an effective translation procedure that, for a given formula $\varphi$, yields a formula $\varphi^\#$, such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

	Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning:
	It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
	Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$.
	Still, we propose a similar transformation.
	A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity and no further processing is needed.

	\begin{theorem}\label{thm:Tseytin1}
		For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
		$\text{ \hspace{1cm}} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
		for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
	\end{theorem}

	The main transformation then proceeds in three steps:
	%
	1) We encode as a first-order sentence that the considered formula holds for every Kripke frame.
	Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory, this step is rather straightforward.
	%
	2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
	%
	3) We can then argue that if there is a counter-model for the resulting formula, then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
	This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke frame.

	Summarizing, we obtain the following result:

	\begin{theorem}
		\label{thm:reduction-propositional}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin1} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
		\begin{itemize}
			\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
			\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occur in $k$.
		\end{itemize}
		Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Furthermore, there is an effective procedure for translating counter-models between the sentences.
	\end{theorem}

	Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
	We can show that this QBF is satisfiable iff there exists a counter-model for our original formula:

	\begin{theorem}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$. There is an effective procedure that produces a QBF $\varphi^Q$ of size $\mathcal O(|\mathcal S|\cdot|\mathcal F_\to| + |\mathcal F_\to|^3)$ with $2\cdot |\mathcal F_\to|-1$ quantifier alternations such that $\varphi$ is intuitionistically valid if and only if $\varphi^Q$ is not satisfiable. For fixed $N\in\mathbb N$, deciding intuitionistic validity for formulas $\bigwedge \mathcal S\to P$ where $\mathcal S$ is as above and $|\mathcal F_\to| = N$ is therefore in $\Sigma_{2N-1}$.
	\end{theorem}

	We then move to the first-order case. Here the construction is a bit more involved, but the underlying approach is the same.
	Again, we first perform a Tseytin-like transformation.
	Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
	The details of this translation are given in Section~\ref{section:tseytin}.
	We then proceed similarly to the propositional case.
	Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame, but on the other hand also proper domain elements.
	We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world.
	The main transformation then again proceeds in three steps:
	Step 1) and 2) are analogous to the propositional case.
	The real difference arises in step 3):
	While we can also establish the existence of a canonic counter-model whose frame only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
	This is not surprising, since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models.
	Therefore, we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics.
	However, since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable.

	Summarizing, we obtain the following result:

	\begin{theorem}
		\label{thm:reduction-first-order-short}
		There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $\varphi$, however, for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$, and $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore, there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
	\end{theorem}
	A more detailed account of Theorem~\ref{thm:reduction-first-order-short} is given in Theorem~\ref{fullFOtranslation}.

	\section{Implementation}

	In this section, we describe our system that implements the translation from the previous section. The rust source code can be found online~\cite{implementation}. We then elaborate our benchmarking process, which utilizes the ILTP problem set~\cite{iltp} and the Vampire theorem prover~\cite{Kov_cs_2013}.

	\begin{figure}[h]\label{results}
		$$\begin{matrix}
			&\text{AGT}&\text{ALG}&\text{COM}&\text{CSR}&\text{GEJ}&\text{GEO}&\text{GPJ}&\text{GRA}\\
			\text{Proven}&5&170&0&0&58&3&4&2\\
			\text{Disproven}&0&0&0&0&0&0&0&0\\
			\text{Timeout}&47&29&3&29&107&74&1&16
		\end{matrix}$$
		$$\begin{matrix}
			&\text{GRP}&\text{HAL}&\text{KRS}&\text{LCL}&\text{MGT}&\text{MSC}&\text{NLP}&\text{NUM}\\
			\text{Proven}&0&0&30&1&9&1&9&2\\
			\text{Disproven}&0&0&0&0&12&0&0&0\\
			\text{Timeout}&5&9&128&3&69&1&237&82
		\end{matrix}$$
		$$\begin{matrix}
			&\text{PLA}&\text{PUZ}&\text{SET}&\text{SWC}&\text{SWV}&\text{SYJ}&\text{SYN}&\text{TOP}\\
			\text{Proven}&0&5&27&1&64&65&54&1\\
			\text{Disproven}&0&1&0&0&2&3&13&0\\
			\text{Timeout}&6&1&297&422&154&108&298&2
		\end{matrix}$$
		\caption{Performance on problem sections of the ILTP set.}
	\end{figure}

	Our implementation performs similar to existing dedicated provers for intuitionistic logic, but falls short of the state-of-the-art. Since there is still a lot of possible optimizations left for our translation, this is a hopeful first sign.

	\section{Translation of counter-models and proofs}

	In the final section, we give an explicit mechanism how to transform the proofs and counter-models for translated formula $\varphi^\#$ into intuitionistic proofs/counter-models of the original formula $\varphi$. The existence of these transformations is clear from the methodology of chapter~\ref{chapter:eiicl}, however, they are hidden in the proofs and here we make them explicit. The two main results are as follows:
	\begin{theorem}
		There exists an effective procedure that converts any classical counter-model for a translated formula $\varphi^\#$ into an intuitionistic counter-model for the original formula $\varphi$.
	\end{theorem}
	\begin{theorem}
		There exists an effective procedure that converts any classical proof of a translated formula $\varphi^\#$ into a intuitionistic proof of the original formula $\varphi$.
	\end{theorem}

	Given the arguments of Chapter~\ref{chapter:eiicl} the translation of counter-models is more or less just a compilation of proofs. On the other hand, the explicit translation of proofs does require some work, since to that point all arguments were semantic and this requires a syntactic viewpoint.

	\chapter{Preliminaries}
	First, let us recapitulate the most important definitions and some facts about them. This will serve to fix our notation as well.

	\section{Syntax}

	\subsection{Propositional logic}

	Let us fix some countably infinite set of propositional variables $A, B, C\dots$.

	\begin{definition}
		A \textit{formula} and \textit{subformula} is defined inductively via the following rules:
		\begin{itemize}
			\item every propositional variable as well as a special symbol $\bot$ is a formula, such a formula is called \textit{atomic} and $A,B,C,\dots$ and $\bot$ are called \textit{atoms}.
			\item if $\varphi$ and $\psi$ are formulas then $\varphi\circ\psi$ is a formula for $\circ\in\{\wedge,\vee,\to\}$.
			\item every formula is a subformula of itself.
			\item for a formula $\chi = \varphi\circ\psi$ every subformula of $\varphi$ and $\psi$ is also a subformula of $\chi$.
		\end{itemize}
	We write $\neg \varphi$ for $\varphi\to \bot$.
	\end{definition}

	\subsection{Predicate logic}

	The analogon for predicate logic is much more involved as we need to define terms.

	\begin{definition}
		A \textit{signature} consists of
		\begin{itemize}
			\item a finite set $S_f$ of function symbols $f_1, \dots, f_n$
			\item a finite set $S_R$ of relation symbols $R_1,\dots, R_m$
			\item  a function ar$: S_f\cup S_R\to \mathbb{N}$ assigning to each symbol its arity (possibly 0). 
		\end{itemize}
	\end{definition}

	\noindent Usually the signature will be left implicit. If explicitly stated we denote it by $$\{f_1/\text{ar}(f_1),\dots,f_n/\text{ar}(f_n), R_1/\text{ar}(R_1).\dots,R_m/\text{ar}(R_m)\}.$$
	Fix a countably infinite collection of free variables $a, b, c\dots$ and bound variables $x, y, z\dots$.
	Note that the following definitions could be simplified by not distinguishing between bound and free variables. However, this has some other drawbacks and in particular when defining a sequent calculus for first-order logic having disjoint sets of bound and free variables is convenient.

	\begin{definition}
		A \textit{semiterm} is defined inductively via the following rules
		\begin{itemize}
			\item each variable is a term
			\item if $t_1,\dots,t_n$ are terms and $f$ a $n$-ary function symbol then $f(t_1, \dots, t_n)$ is a term.
		\end{itemize}
		A \textit{term} is a semiterm in which each occurring variable is a free variable. Every term is a semiterm. Analogously to the definition of subformulas we can define \textit{subsemiterms} and \textit{subterms}.
	\end{definition}

	\begin{definition}
		A \textit{Substitution} is a function $\sigma$ from free variables to semiterms. For a semiterm $t$ we define $t\sigma$ as follows:
		\begin{itemize}
			\item $x\sigma = x$ if $x$ is bound.
			\item $a\sigma = \sigma(a)$ if $a$ is free.
			\item $f(t_1,\dots,t_n)\sigma = f(t_1\sigma,\dots,t_n\sigma)$.
		\end{itemize}
	\end{definition}

	\begin{definition}
		A \textit{formula} is defined inductively via the following rules
		\begin{itemize}
			\item $\bot$ is a formula.
			\item if $t_1,\dots,t_n$ are terms and $R$ and an $n$-ary relation symbol then $R(t_1, \dots, t_n)$ is a formula.
			\item if $s, t$ are terms then $s = t$ is a formula.
		\end{itemize}	
		Formulas formed by the above rules are called \textit{atomic}.
		\begin{itemize}
			\item If $\varphi$ and $\psi$ are formulas then $\varphi\circ\psi$ is a formula for $\circ\in\{\wedge,\vee,\to\}$.
			\item If $\varphi$ is a formula and $a$ is a free variable occurring in $\varphi$ and $x$ is a bound variable not occurring in $\varphi$ then $Qx(\varphi[x/a])$ is a formula for $Q\in\{\exists,\forall\}$ where the \textit{substitution} $\varphi\sigma$ is defined inductively via
			\begin{itemize}
				\item $\bot\sigma = \bot$
				\item $R(t_1,\dots,t_n)\sigma = R(t_1\sigma,\dots, t_n\sigma)$
				\item $(s = t)\sigma = (s\sigma = t\sigma)$
				\item $(\varphi\circ\psi)\sigma = \varphi\sigma\circ\psi\sigma$ for $\circ\in\{\wedge,\vee,\to\}$.
				\item $(Qy\varphi)\sigma = Qy(\varphi\sigma)$ for $Q\in\{\exists,\forall\}$.
			\end{itemize}
			\item A formula containing no free variables is a \textit{sentence}.
		\end{itemize}
	\end{definition}

	\section{Semantics}

	Classical and intuitionistic semantics give (different) meaning to formulas.

	\subsection{Propositional Logic}
	\subsubsection{Classical Semantics}

	\begin{definition}
		A \textit{valuation} $v$ is a function that assigns to each propositional variable $A$ a \textit{truth value} $v(A)\in\{0, 1\}$. We define a model relation $\models$ between valuations $v$ and formulas:
		\begin{itemize}
			\item $v\not\models \bot$
			\item $v\models A$ iff $v(A) = 1$ for each propositional variable $A$.
			\item $v\models \varphi\wedge\psi$ iff $v\models\varphi$ and $v\models\psi$.
			\item $v\models\varphi\vee\psi$ iff $v\models\varphi$ or $v\models\psi$.
			\item $v\models\varphi\to \psi$ iff $v\not\models\varphi$ or $v\models\psi$.
		\end{itemize}
		$v$ is a \textit{model} for $\varphi$ if $v\models\varphi$. $\varphi$ is \textit{satisfiable} if there exists a valuation which is a model. It is \textit{valid} if every valuation is a model, and we write $\models \varphi$. We denote the set of valid formulas with CPC. We can furthermore define an entailment relation between sets of clauses, that is $\mathcal S\models_C \mathcal T$ if for every valuation $v$ with $v\models\bigwedge S$ we have $v\models\bigvee \mathcal T$.
	\end{definition}

	One notable property of classical logic is that a formula $\varphi$ is valid iff its negation, i.e. $\neg\varphi$ is not satisfiable. The same does not hold true for intuitionistic logic, as we shall see.

	\subsubsection{Intuitionistic Semantics}

	\begin{definition}
		A \textit{Kripke structure} $\mathcal K = (W, (v_w)_{w\in W})$ consists of a partially ordered set $W$ of nodes an a family of valuations $(v_w)_{w\in W}$ such that for $u\leq w$ we have $v_u(A)\leq v_u(A)$ for every propositional variable $A$, this is called the \textit{persistency condition}. In this case, we define a model relation between worlds $u$ and formulas. It is slightly different in the case of implications, i.e. have
		\begin{itemize}
			\item $u\not\models \bot$
			\item $u\models A$ iff $v_u(A) = 1$ for each propositional variable $A$.
			\item $u\models \varphi\wedge\psi$ iff $v\models\varphi$ and $v\models\psi$.
			\item $u\models\varphi\vee\psi$ iff $v\models\varphi$ or $v\models\psi$.
			\item $u\models\varphi\to \psi$ if for all $w\geq u$ we have $w\not\models\varphi$ or $w\models\psi$.
		\end{itemize}
		We say that $\varphi$ is satisfied at a node $u$ if $u\models\varphi$. If a formula is satisfied at every node, then $K$ is a \textit{model} for $\varphi$, and we write $K\models \varphi$. $\varphi$ is satisfiable if there exists a Kripke structure, which is a model. It is \textit{valid} if every Kripke structure is a model and we write $\models \varphi$. We denote the set of valid formulas with IPC. As before, we define an entailment relation $\models_I$ between sets of formulas.
	\end{definition}
	There are many classical tautologies which are not intuitionistically valid
	\begin{example}\label{LEMcounterexample}
		Consider the famous law of excluded middle $A\vee\neg A$. As a counter model consider a Kripke structure with two nodes $u\leq w$ and $v_u(A) = 0, v_w(A) = 1$. Then $u\not\models A$ and $u\not\models \neg A$, i.e. $u\not\models A\vee\neg A$. 
	\end{example}
	The following is a famous and important result due to Glivenko~\cite{glivenko1929}

	\begin{theorem}[Glivenko's theorem]
		A propositional formula $\varphi$ is classically valid iff $\neg\neg\varphi$ is intuitionistically valid.
	\end{theorem}
	\noindent Note however that this difference between classical and intuitionistic logic is not reflected in satisfiability, only in validity:
	\begin{lemma}\label{propsat}
		A propositional formula $\varphi$ is intuitionistically satisfiable if and only if it is classically satisfiable.
	\end{lemma}

	\begin{proof}
		Every classical model can be viewed as a Kripke model with a single node. Therefore, classical satisfiability trivially implies intuitionistic satisfiability.	On the other hand, in every Kripke structure that is a model every node is a classical model so intuitionistic satisfiability implies intuitionistic satisfiability.
	\end{proof}

	In classical logic, a proven strategy for showing that a formula valid is to consider its negation and show that it is not satisfiable. Due to the previous lemma, this is insufficient for intuitionistic logic. There are formulas like $A\vee\neg A$ which are not an intuitionistic tautology, while their negation is not satisfiable either. This serves to motivate the consideration of equivalidity in the following chapters. In the context of classical logic it can be fully explained via equisatisfiability and is therefore neglected, but for intuitionistic logic it is more fundamental.

	Finally, let us note the complexities of these various decision problems, which play a fundamental role in all of theoretical computer science.

	\begin{theorem}
		Satisfiability for classical and intuitionistic propositional logic is \verb+NP+-complete.
	\end{theorem}

	For classical logic this is the famous Cook-Levin theorem first published in Cook's seminal paper~\cite{cook1971complexity}. Of course, we have seen that the intuitionistic case is the same decision problem.

	\begin{corollary}
		Validity for classical logic is \verb+co-NP+-complete
	\end{corollary}

	This is clear, since validity is just the complementary problem to satisfiability of the negated formula. However, in intuitionistic logic unfortunately validity is really harder than satisfiability as first shown by Statman~\cite{statman1979intuitionistic}.

	\begin{theorem}
		Validity for intuitionistic logic in \verb+PSPACE+-complete.
	\end{theorem}

	One thing to note is that in contrast to predicate logic, all these problems are decidable.

	\subsection{Predicate Logic}

	We now give account of the more involved semantics of first-order logic

	\subsubsection{Classical Semantics}

	\begin{definition}
		Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the domain of $\mathcal{M}$, and a function $I$ that assigns
		\begin{itemize}
			\item to each $n$-ary function symbol $f$ a $n$-ary function $f^I: M^n\to M$.
			\item to each $n$-ary relation symbol $R$ a $n$-ary relation $R^I\subseteq M^n$.
		\end{itemize}
		A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
			m&\text{if $b=a$}\\
			v(b)&\text{otherwise}
		\end{cases}$$
		Then terms are interpreted as follows:
		\begin{itemize}
			\item $a^{I, v} = v(a)$ for each free variable $a$.
			\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$.
		\end{itemize}
		We define a model relation between pairs $\mathcal M, v$ and formulas $\varphi$ as follows
		\begin{itemize}
			\item $\mathcal M, v\not\models\bot$.
			\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$.
			\item $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
			\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$.
			\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
			\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
			\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
			\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
		\end{itemize}
		$\varphi$ is satisfiable if for some $\mathcal{M}, v$ we have  $\mathcal M, v\models\varphi$. It is valid if for all  $\mathcal M, v$ we have  $\mathcal M, v\models\varphi$. We write $\mathcal{M}\models\varphi$ if for every $v$ we have  $\mathcal M, v\models\varphi$. We denote the set of valid formulas with CQC. As before define an entailment relation $\models_C$ between sets of formulas.
	\end{definition}

	\subsubsection{Intuitionistic Semantics}

	\begin{definition}
		A $\Sigma$-\textit{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$ and a family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$
		\begin{itemize}
			\item $M_u\subseteq M_w$.
			\item $f^{I_w}|_{M_u} = f^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$.}
			\item $R^{I_w}|_{M_u} = R^{I_u}$.
		\end{itemize}
		Then we define a model relation between worlds $u\in W$ and variable assignments $v$ (targeting $M_u$) and formulas $\varphi$ as follows:
		\begin{itemize}
			\item $u, v\not\models\bot$
			\item $u, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I_u, v},\dots,t_n^{I_u, v})\in R^{I_u}$.
			\item $u, v\models s = t$ iff $s^{I_u, v} = t^{I_u, v}$.
			\item $u, v\models \varphi\wedge \psi$ iff $u, v\models\varphi$ and $u, v\models\psi$.
			\item $u, v\models \varphi\vee\psi$ iff $u, v\models\varphi$ or $u, v\models\psi$.
			\item $u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $w, v\not\models\varphi$ or $w, v\models\psi$.
			\item $u, v\models\exists x\varphi$ iff there exists $m\in M_u$ such that $u, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
			\item $u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $w, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
		\end{itemize}
		We say that $\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$ and write $\mathcal K\models\varphi$. As always, a formula is satisfiable if it is satisfied by some Kripke structure, and it is valid if it is satisfied in every Kripke structure. We denote the set of valid formulas with IQC.  As before, define an entailment relation $\models_I$ between sets of formulas.
	\end{definition}
	In addition to the propositional tautologies, there are now sentences involving quantifiers, which are classically valid but not intuitionistically.
	Consider for instance the classical tautology $\varphi$ $$\neg\forall x A(x)\to \exists x \neg A(x)$$We exhibit an intuitionistic counter model. Consider the Kripke structure with two nodes $u\leq w$ with $M_u = \{0\}, M_w = \{0, 1\}$ and $A^{I_u} = A^{I_w} = \{0\}$. Then $u, v\not\models \varphi$: First note that $w, v\not\models\forall xA(x)$ and $u, v\not\models\forall xA(x)$ for any valuation $v$ since $w, v[1/a]\not\models A(a)$, i.e. $u, v\models\neg\forall xA(x)$ and also $w, v\models\neg\forall xA(x)$. Furthermore, $u, v[0/a]\not\models \neg A(a)$ and $w, v[0/a]\not\models \neg A(a)$ so $u, v\not\models \exists x\neg A(x)$ for any valuation $v$ (note however that this is not true at $w$!). Therefore, $u, v\not\models\varphi$.

	Intuitively, the problem is that the existential quantifier at $u$ does not ``see'' the elements at $w$, i.e. it must "choose" between the elements at $u$. As in the propositional case, there are well-known negative translations of classical logic into intuitionistic logic going back to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit}.

	\begin{definition}
		For a predicate formula $\varphi$ define inductively $\varphi^N$ as follows:
		\begin{itemize}
			\item $\varphi^N = \varphi$ if $\varphi$ is atomic
			\item $(\varphi\wedge\psi)^N = \varphi^N\wedge\psi^N$
			\item $(\varphi\vee\psi)^N = \neg(\neg\varphi^N\wedge\neg\varphi^N)$
			\item $(\varphi\to\psi)^N = \varphi^N\to\psi^N$
			\item $(\forall x\varphi)^N = \forall x\varphi^N$
			\item $(\exists x\varphi)^N = \neg\forall x\neg\varphi^N$
		\end{itemize}
	\end{definition}
	\begin{theorem}
		A $\Sigma$-formula $\varphi$ is classically valid if and only if $\varphi^N$ is intuitionistically valid.
	\end{theorem}

	But again this is not reflected in satisfiability, i.e. with a completely analogous argument to~\ref{propsat} we obtain

	\begin{lemma}
		Any $\Sigma$-formula $\varphi$ is classically satisfiable if and only if it is intuitionistically satisfiable.
	\end{lemma}

	\section{Skolemization and Herbrandization}

>>>>

\documentclass{article}
\usepackage[english]{babel}
\usepackage{setspace}

\title{Master's Thesis Summary}
\author{Alexander Pluska}
\begin{document}

	\maketitle

	Constructive mathematics refers to a flavour of mathematics in which the existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction.
	The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the fact that the law of excluded middle $A\vee\neg A$ and the double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.
	Besides philosophical considerations, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation for studying constructive mathematics from the perspective of computer science in that proofs in  intuitionistic logic directly correspond to computer programs --- as expressed in the Curry-Howard correspondence~\cite{howard1980formulae}.

@@

	\begingroup
	\setstretch{0.8}
	\bibliographystyle{plain}
	\bibliography{EIICL}
	\endgroup
\end{document}
